Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
Pushing.  Edits and edits accepted for Sec 1.
Couldn't get two footnotes to compile.

1.  Changed 'aid' to help.  Does 'to aid an individual decide the correct approach?' make grammatical sense?  I may be overthinking.

2.  'extend the user toolkit' - wanted to remove the 'give' part.

The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process \ben{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}. \footnote{\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}}

With the many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021}, this field has a lot of devices to aid an individual practitioner decide the correct approach.  For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.

At the same time, analysis packages often only address a specific portion of the analysis pipeline\ben{\sout{, for instance focusing on certain types of models.}} \ben{These m}odeling tools\ben{\sout{, which}} usually require learning package-specific syntax\ben{\sout{,}} and often don't provide easy ways to compare and assess their models on new data. Moreover, exploring\ben{, \sout{and}} modeling \ben{and comparing} epidemics require transforming and \textit{tidying} data in different ways. To fill these gaps, we present our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic and epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.

<!-- # old draft
The recent (and currently on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g. ...][]{}. The COVID-19 pandemic highlights how all three goals are important both separately and taken as a whole.  Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.  Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks. \shannon{really like this paragraph}

The current epidemic of COVID-19 also highlights that infectious disease  models are only one piece of the overall analysis pipeline. University based resources like John Hopkins' and government numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and visualization can be a important first step in the process (multiple \cite{}?).} \shannon{add comment about NYT or something} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individual / agent level, network structure or just aggregate flows (review paper \cite{}?). All placing individuals / proportions of the populations into different states (e.g. suspectible, exposed, infected, recovered, etc.). With all these models, review and comparison papers in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center helps the individual practitioner decide the correct approach. \shannon{along with expertise from healthcare professionals...}

At the same time, analysis packages often only address a portion of the analysis pipeline. Modeling tools often don't provide easy ways to compare and assess their models on new data. Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways. To fill these gaps, we present our our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline. 
-->

@@

[One of \pkg{EpiCompare}’s main contribution to] comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments. Epidemics, despite being defined a a process that evolves overt time, often needs to be compared in a way not constrained to initial times or time scales to understand the processes at play.] [With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible.]\footnote{\ben{[Ben says: brought this sentence from a later paragraph as it’s much more introductory.]}} Compared to time-dependent comparison tools for state-space modeling,  [time-invariant analysis can make it easier to compare state-space \ben{\sout{models} epidemic representations} in a more global, holistic fashion.] Many time-dependent comparison tools only [examine the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.] [These \ben{time-dependent approaches can} reduce the amount of \ben{\sout{connections that can be seen} insights that can be drawn}, similar to \ben{examining} projections of a multidimensional distribution onto a single axis \ben{one} at a time. Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens. \ben{At the same time, t}he goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.]

This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a \ben{\sout{thorough}} demonstrating of the tools through a detailed example of a full data analysis pipeline.

# Time-invariant analysis

>>>>

The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process, as has been seen in the current COVID-19 \shannon{\sout{epi}}\shannon{pan}demic \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}. \footnote{\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}\shannon{Seems fine to me now}}

With \shannon{\sout{the}} many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} \shannon{available}, this field has a \shannon{\sout{lot of}} \shannon{number of} devices to \shannon{\sout{aid}} \shannon{help}  an individual practitioner decide the correct approach. For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.

At the same time, analysis packages often only address a specific portion of the analysis pipeline. These modeling tools usually require learning package-specific syntax and often don't provide easy ways to compare and assess their models on new data. Moreover, exploring modeling and comparing epidemics require transforming and \textit{tidying} data in different ways. To fill these gaps, we present our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic and epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.

@@

One of \pkg{EpiCompare}’s main contribution to comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments. Epidemics, despite being defined as a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales \shannon{in order}\footnote{\shannon{wanted a clearer signal that this is a bridge to a new thought}}to understand the processes at play. With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible. Compared to time-dependent comparison tools for state-space modeling,  time-invariant analysis can make it easier to compare state-space epidemic representations in a more global, holistic fashion. Many time-dependent comparison tools only examine the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion. These time-dependent approaches can reduce the amount of connections that can be seen and insights that can be drawn}, similar to examining projections of a multidimensional distribution onto a single axis, one at a time. Tools in \pkg{EpiCompare} \shannon{\sout{give the user the ability to extend their} \shannon{extend the user}} toolkit to evaluate epidemics within a time-invariant lens. At the same time, the goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a  demonstration of the tools through a detailed example\shannon{\sout{ of a full data analysis pipeline.}} \shannon{from start to finish of the data analysis pipeline}\footnote{\shannon{maybe clearer? up to you}}

# Time-invariant analysis {#sec:time-invariant}

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper2.tex
Pushing.  Edits and edits accepted for Sec 1.
Couldn't get two footnotes to compile.

1.  Changed 'aid' to help.  Does 'to aid an individual decide the correct approach?' make grammatical sense?  I may be overthinking.

2.  'extend the user toolkit' - wanted to remove the 'give' part.

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
and visualizations from universities, many branches and levels of
government, and news organizations are an important first step of the
process
\textcolor{violet}{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}}

With the many visualization and exploratory tools, models and modeling
paradigms, and reviews and comparisons in the literature and through the
MIDAS (Models of Infectious Disease Agent Study) network
\citep{midasnetwork2021}, this field has a lot of devices to aid an
individual practitioner decide the correct approach. For example,
\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
\pkg{pomp} have all made significant steps in standardizing the flow of
the data analysis pipeline for epidemic modeling through digitizing data
sets, making accessible statistical models, and providing a plethora of
educational material for both coding novices and experts alike
\citep{surveillance2017,Jenness2018,King2016}.

At the same time, analysis packages often only address a specific
portion of the analysis
pipeline\textcolor{violet}{\sout{, for instance focusing on certain types of models.}}
\textcolor{violet}{These m}odeling
tools\textcolor{violet}{\sout{, which}} usually require learning
package-specific syntax\textcolor{violet}{\sout{,}} and often don't
provide easy ways to compare and assess their models on new data.
Moreover, exploring\textcolor{violet}{, \sout{and}} modeling
\textcolor{violet}{and comparing} epidemics require transforming and
\textit{tidying} data in different ways. To fill these gaps, we present
our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
is to aid and advance research in the area of comparison and assessment
of epidemic and epidemiological models. In Figure \ref{fig:pipeline}, we
illustrate the data analysis pipeline of infectious diseases as 1) data
pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
simulating, 4) post-processing, and 5) comparison and assessment; where
each previous part of the pipeline influences the next. \pkg{EpiCompare}
provides tools to aids practitioners in all areas of this pipeline.

@@

{[}One of \pkg{EpiCompare}'s main contribution to{]} comparison and
assessment of epidemics is through tools that provide
\textit{time-invariant} assessments. Epidemics, despite being defined a
a process that evolves overt time, often needs to be compared in a way
not constrained to initial times or time scales to understand the
processes at play.{]} {[}With time-invariant analysis, comparing
decades-long outbreaks of HIV in the US to a 10 day outbreak of
norovirus on a cruise ship is
possible.{]}\footnote{\textcolor{violet}{[Ben says: brought this sentence from a later paragraph as it’s much more introductory.]}}
Compared to time-dependent comparison tools for state-space modeling,
{[}time-invariant analysis can make it easier to compare state-space
\textcolor{violet}{\sout{models} epidemic representations} in a more
global, holistic fashion.{]} Many time-dependent comparison tools only
{[}examine the proportion of individuals in each state (at a given time)
in a piece-wise / marginal fashion.{]} {[}These
\textcolor{violet}{time-dependent approaches can} reduce the amount of
\textcolor{violet}{\sout{connections that can be seen} insights that can be drawn},
similar to \textcolor{violet}{examining} projections of a
multidimensional distribution onto a single axis \textcolor{violet}{one}
at a time. Tools in \pkg{EpiCompare} give the user the ability to extend
their toolkit to evaluate epidemics within a time-invariant lens.
\textcolor{violet}{At the same time, t}he goal of \pkg{EpiCompare} is
not to supplant existing infectious disease modeling tools and software
but, rather, is a concerted effort to create standard and fair
comparisons among models developed for disease outbreaks and outbreak
data.{]}

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a \textcolor{violet}{\sout{thorough}}
demonstrating of the tools through a detailed example of a full data
analysis pipeline.

\hypertarget{time-invariant-analysis}{%
\section{Time-invariant analysis}\label{time-invariant-analysis}}

>>>>

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
and visualizations from universities, many branches and levels of
government, and news organizations are an important first step of the
process, as has been seen in the current COVID-19
\textcolor{orange}{\sout{epi}}\textcolor{orange}{pan}demic
\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}\textcolor{orange}{Seems fine to me now}}

With \textcolor{orange}{\sout{the}} many visualization and exploratory
tools, models and modeling paradigms, and reviews and comparisons in the
literature and through the MIDAS (Models of Infectious Disease Agent
Study) network \citep{midasnetwork2021} \textcolor{orange}{available},
this field has a \textcolor{orange}{\sout{lot of}}
\textcolor{orange}{number of} devices to \textcolor{orange}{\sout{aid}}
\textcolor{orange}{help} an individual practitioner decide the correct
approach. For example, \proglang{R} packages such as \pkg{surveillance},
\pkg{EpiModel}, and \pkg{pomp} have all made significant steps in
standardizing the flow of the data analysis pipeline for epidemic
modeling through digitizing data sets, making accessible statistical
models, and providing a plethora of educational material for both coding
novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.

At the same time, analysis packages often only address a specific
portion of the analysis pipeline. These modeling tools usually require
learning package-specific syntax and often don't provide easy ways to
compare and assess their models on new data. Moreover, exploring
modeling and comparing epidemics require transforming and
\textit{tidying} data in different ways. To fill these gaps, we present
our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
is to aid and advance research in the area of comparison and assessment
of epidemic and epidemiological models. In Figure \ref{fig:pipeline}, we
illustrate the data analysis pipeline of infectious diseases as 1) data
pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
simulating, 4) post-processing, and 5) comparison and assessment; where
each previous part of the pipeline influences the next. \pkg{EpiCompare}
provides tools to aids practitioners in all areas of this pipeline.

@@

One of \pkg{EpiCompare}'s main contribution to comparison and assessment
of epidemics is through tools that provide \textit{time-invariant}
assessments. Epidemics, despite being defined as a process that evolves
over time, often need to be compared in a way not constrained to initial
times or time scales
\textcolor{orange}{in order}\footnote{\textcolor{orange}{wanted a clearer signal that this is a bridge to a new thought}}to
understand the processes at play. With time-invariant analysis,
comparing decades-long outbreaks of HIV in the US to a 10 day outbreak
of norovirus on a cruise ship is possible. Compared to time-dependent
comparison tools for state-space modeling, time-invariant analysis can
make it easier to compare state-space epidemic representations in a more
global, holistic fashion. Many time-dependent comparison tools only
examine the proportion of individuals in each state (at a given time) in
a piece-wise / marginal fashion. These time-dependent approaches can
reduce the amount of connections that can be seen and insights that can
be drawn\}, similar to examining projections of a multidimensional
distribution onto a single axis, one at a time. Tools in
\pkg{EpiCompare}
\textcolor{orange}{\sout{give the user the ability to extend their} \textcolor{orange}{extend the user}}
toolkit to evaluate epidemics within a time-invariant lens. At the same
time, the goal of \pkg{EpiCompare} is not to supplant existing
infectious disease modeling tools and software but, rather, is a
concerted effort to create standard and fair comparisons among models
developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a demonstration of the tools through a
detailed
example\textcolor{orange}{\sout{ of a full data analysis pipeline.}}
\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer? up to you}}

\hypertarget{sec:time-invariant}{%
\section{Time-invariant analysis}\label{sec:time-invariant}}
